---
title: "Reproducible Research Peer Review 1"
author: "Jim Maloney"
date: "October 18, 2015"
output: html_document
---

##Introduction

This is an R markdown document for the first project in the Reproducible Research class.
The project takes fitbit / daily step data recorded at 5 minute intervals over a roughly 2 month period. 
The introductory step is to read in the data file.

```{r}  
myfile <- read.csv ('activity.csv',header = TRUE, sep =',')
library (dplyr,plyr)
```

-------------------------------------------

##Part I - Daily Activity Levels
The first part of the assignment is to aggregate and compile data around daily step data.  
This includes graphing a histogram of the number of steps traveled each day
as well as a summary of mean and median daily steps taken. 


```{r}
dailysteps <- group_by(myfile,date)
totaldailysteps <- summarize (dailysteps, Daily = sum(steps, na.rm = TRUE))
totaldailysteps <- as.data.frame(totaldailysteps)

## create a histogram of the daily step activity
quicklook <- hist(totaldailysteps$Daily)
summary (totaldailysteps$Daily)
meandailysteps <- mean(totaldailysteps$Daily)
mediandailysteps <- median(totaldailysteps$Daily)
```

**So from above, we can see that the mean daily steps are 9354 and the median daily steps are 10395.**

---------------------------------------------

##Part II - IntraDay Activity Levels

The second section of the assignment was to look at intraday step patterns. 
This means we need to group by interval rather than by day and then run similar calculations as per Part I. above.
The first output will be a time series graph that shows avg steps taken by 5 min interval throughout the day.

```{r}
intervalsteps <- group_by (myfile,interval)
avgintervalsteps <- as.data.frame (summarize (intervalsteps, AvgInterval = mean(steps, na.rm = TRUE)))

# graph a time series analysis to get a feel for high/low interval periods
intervalstepgraph <- plot (avgintervalsteps$interval,avgintervalsteps$AvgInterval, type = 'l')
```

The next output is to determine the most active interval of the day.  Note that the intervals run in 5 minute increments for each
hour of the day. 

```{r}
summary(avgintervalsteps$AvgInterval)

mostactiveinterval <- max(avgintervalsteps$AvgInterval)
attach(avgintervalsteps)
topinterval <- avgintervalsteps [AvgInterval == max(AvgInterval), ]
paste ('The busiest time is at ', topinterval$interval, 'with an average of ',topinterval$AvgInterval, ' steps taken.')
```

**So from above, we can see that the busiest time is 8:35 am and that on average 206 steps are taken.**

---------------------------------------------

##Part III - Imputing Missing Values

The next part of the exercise is to impute missing values and see how the results from I. and II. change as a result.
The first thing to do is to determine how many missing items there are.  This will be done through a simple count of NAs.

```{r}

nostepinfo <- sum(is.na(myfile$steps))
paste ('There are ', nostepinfo, ' missing entries in the dataset.')

```

** This shows 2304 missing entries.**

The next step is to replace missing data elements w/ proxies.  
While it's not clear what the best option is, I prefer to take the average of the 5 minute interval for the available dates
and fill in the dataset that way. I believe the potential error here would be less, esp since it appears that intervals
have higher variance than days.

```{r}
#imputed dataset - first copy from original data (myfile)
impdataset <- myfile
# will first append the average interval step data to the original data file.
# since every day has the same intervals, this will copy over cleanly
imputedataset <- cbind(impdataset,avgintervalsteps$AvgInterval)

# will rename the last column avgstepint (average steps per interval)
names (imputedataset) [4] <- c('avgstepint')
# check to make sure all looks good
str(imputedataset)
# first NAs end around row 288.
imputedataset[285:335,]

# this is the key step where we update the step data from NA to the avg for that interval.  if not NA, then we leave it as is.
imputedataset$avgstepint <- ifelse(is.na(imputedataset$steps),imputedataset$avgstepint,imputedataset$steps)
# check - now should see last column change over from avg to actual step data
imputedataset [285:335,]

```

Now having created the imputed data set, we will rerun the histogram and recalculate the means/medians etc.

```{r}
# first the histogram - need to group by day

imputedailysteps <- group_by(imputedataset,date)
imputedailysteptotal <- summarize (imputedailysteps, Daily= sum(avgstepint))
imputedailysteptotal <- as.data.frame(imputedailysteptotal)
imputequicklook <- hist(imputedailysteptotal$Daily)

```

Then to caclualte the means and medians.

```{r}
summary(imputedailysteptotal$Daily)
meanimputed <- mean(imputedailysteptotal$Daily)
medianimputed <- median(imputedailysteptotal$Daily)
```

**From the above we can see that both the mean and the median daily steps are now 10,770.**

Then to calculate the analysis of difference between observed and imputed data.

```{r}

ImputeLessObservedDailyMean <- meanimputed- meandailysteps
ImputeLessObservedDailyMedian <- medianimputed -mediandailysteps
paste('For daily steps taken, the imputed mean is greater than the observed mean by ',ImputeLessObservedDailyMean)
paste('For daily steps taken, the imputed median is greater than the observed median by ', ImputeLessObservedDailyMedian)
```

**From this we see the mean has increased more than the median.**

##Part IV. Comparison of Weekday vs. Weekend Daily Activity

The last part of the analysis is to compare average daily steps taken on weekdays vs. weekends, and then see if there is
a significant difference in intra-day patterns as well.  Will use the imputed data set to produce the charts needed here.

The first chart will be a comparison of daily activity between weekdays and weekends.  
Need to identify and sort data by weekday/weekend.

```{r}
# first need to convert date data from factor to date
imputedataset$date <- as.Date(imputedataset$date)
# then need to append a column that tells first what day of the week it is
# then create another column appending whether the day is a weekday or weekend 
imputedataset$day <- weekdays(imputedataset$date, abbreviate = TRUE)
imputedataset$daytype <- ifelse(imputedataset$day %in% c("Sat","Sun"),'Day','End')
# now make a time series graph of each type
# first step is to group by interval and day type
weekdaydata <- filter(imputedataset, imputedataset$daytype == 'Day')
weekenddata <- filter(imputedataset,imputedataset$daytype ==  'End')

#now determine averages for weekday data
wkdayintervalsteps <- group_by (weekdaydata,interval)
avgwkdayintervalsteps <- as.data.frame (summarize (wkdayintervalsteps, AvgInterval = mean(avgstepint)))
# graph a time series analysis to get a feel for high/low interval periods
wkdayintervalstepgraph <- plot (avgwkdayintervalsteps$interval,avgwkdayintervalsteps$AvgInterval, type = 'l')
# this shows a graphy where all the activity takes place at 8.30 am with a smallder spike at 5 pm

# now run for weekend data
wkendintervalsteps <- group_by (weekenddata,interval)
avgwkendintervalsteps <- as.data.frame (summarize (wkendintervalsteps, AvgInterval = mean(avgstepint)))
# graph a time series analysis to get a feel for high/low interval periods
wkendintervalstepgraph <- plot (avgwkendintervalsteps$interval,avgwkendintervalsteps$AvgInterval, type = 'l')
# this graph shows many more spikes through out the day - starting a bit later, but much higher activity - fitting what
# intuition would tell us.

